Estimating Information Transfer in Hard Parameter Sharing via Precise Bias-variance Asymptotics

### 2021-07-16 revision_journal/exp.tex
finish sec 4 figures

\begin{figure}[!t]
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_covariate_shift.eps}
		\caption{Covariate shift}
		\label{fig_sec5_covariate}
	\end{subfigure}\hfill%
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_model_shift.eps}
		\caption{Model shift}
		\label{fig_sec5_model}
	\end{subfigure}\hfill%
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_cov_model_shift.eps}
		\caption{Covariate and model shift}
		\label{fig_sec5_cov_model}
	\end{subfigure}
	\caption{We compare the empirical excess risk of HPS and several natural transfer learning estimators. We find that HPS achieves the lowest excess risk for predicting task two under (either or both) covariate and model shift. This simulation uses $p = 100, n_1 = n_2 = 200, \sigma = 1/2$. Figure \ref{fig_sec5_cov_model} uses $\lambda = 4$ for generating covariate-shifted data.}
	\label{fig_sec51}
\end{figure}

@@

Figure \ref{fig_sec51} shows the result under (either or both) covariate and model shift.
Figure \ref{fig_sec5_covariate} uses the same data generating process as Figure \ref{fig_sec31}.
Figure \ref{fig_sec5_model} uses the same data generating process as Figure \ref{fig_sec3_model_shift}.
Figure \ref{fig_sec5_cov_model} further combines the covariate-shifted data and the model shift setting.
For reach run, we average the result over 100 random seeds because of high variance from using only a limited number of samples.
We find that HPS outperforms all the other estimators in this simulation (the result for the TL estimator is worse than STL and is omitted from the figure).

>>>>

\begin{figure}[!t]
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_covariate_shift.eps}
		\caption{Covariate shift}
		\label{fig_sec5_covariate}
	\end{subfigure}\hfill%
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_model_shift.eps}
		\caption{Model shift}
		\label{fig_sec5_model}
	\end{subfigure}\hfill%
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/compare_risk_cov_model_shift.eps}
		\caption{Covariate and model shift}
		\label{fig_sec5_cov_model}
	\end{subfigure}
	\caption{We compare the empirical excess risk of HPS and several natural transfer learning estimators. We find that HPS achieves the lowest excess risk for predicting task two under (either or both) covariate and model shift. This simulation uses $p = 100, n_1 = n_2 = 200, \sigma = 1/2$. Figure \ref{fig_sec5_cov_model} uses $\lambda = 4$ for generating covariate-shifted features.}
	\label{fig_sec51}
\end{figure}

@@

Figure \ref{fig_sec51} shows the result under (either or both) covariate and model shifts.
Figure \ref{fig_sec5_covariate} uses the same data generating process as Figure \ref{fig_sec31}.
Figure \ref{fig_sec5_model} uses the same data generating process as Figure \ref{fig_sec3_model_shift}.
Figure \ref{fig_sec5_cov_model} combines both by generating covariate-shifted features and different linear models for each dataset.
For reach run, we average the result over 100 random seeds because of high variance from the small sample sizes.
We find that HPS outperforms all the other estimators in this simulation (the result for the TL estimator is worse than STL and is omitted from the figure).

### 2021-07-16 revision_journal/results.tex
finish sec 4 figures

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{-1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, $\frac{p}{n_1+n_2} \in (2 {\tau}, \frac 2 {\tau})$).
Our result can be viewed as a generalization of the classical result in Lemma \ref{fact_tr} to the \textit{covariate shift} setting.
In particular, Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ are all equal to one, and $a = 1$. Then, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above together with the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the above solutions back to the variance limit, we thus find that the limit is equal to
\begin{align}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since the singular values of $M$ are equal to $1$} \nonumber \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying $\alpha_1$ and $\alpha_2$ from above} \nonumber \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p}, \label{g_id}
\end{align}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the one in Lemma \ref{fact_tr} for this special case.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.

@@

Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following result regarding $g(\id_{p \times p})$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2 n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write down the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        (\alpha_1 M^{\top} M + \alpha_2)^{-1} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \ge \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
As a corollary of Claim \ref{prop_covariate}, when $n_1$ is much larger than $n_2 / \tau^2$, the excess risk of HPS is approximately optimal in the set of all possible $M\in\cS$ when there is no covariate shift between dataset one and two.

@@

\paragraph{Illustrative examples.}
While the estimates from Theorem \ref{cor_MTL_loss} are generally complex, we describe a simplified result for the random-effect model (cf. Section \ref{sec_data}).
Under the assumptions stated in Proposition \ref{lem_hat_v}, $\hat{a}$ is approximately equal to one.
Then, we can calculate that
\[ \cL_1(1) = \frac{p}{n_1 + n_2 - p} \text{ and } \cL_2(1) = \frac{n_1^2(n_1 + n_2 - p) + p n_1 n_2}{(n_1 + n_2)^2 (n_1 + n_2 - p)}, \text{ in which } \kappa(1) = \frac{n_2^2}{(n_1 + n_2)^2 - n_1 p}. \]
Thus, using that $\|\beta^{(1)}-\beta^{(2)}\|^2 = (2+\oo(1))\mu^2$ w.h.p., we conclude that for the random-effect model, the excess risk $L(\hat{\beta}_2^{\MTL}$ is approximately equal to
$$ g(n_1) \define \sigma^2 \cal L_1(1) +   2\mu^2 \cal L_2(1)= \frac{p\sigma^2}{n_1+n_2-p} +  2\mu^2 \cdot \frac{n_1^2 (n_1+n_2-p)+p n_1n_2}{(n_1+n_2)^2(n_1+n_2-p)}$$
plus lower order terms (that vanishes to zero as $p$ goes to infinity).
Next, we provide a detailed analysis of when combining task one using HPS transfers positively to task two, depending on sample sizes $n_1, n_2$, and model shift $\mu$.

@@

As a remark, when $p < n_2 < 3p$, similar results can be derived using the above arguments (details omitted).
Figure \ref{fig_sec3_model_shift} illustrates Claim \ref{claim_model_shift} for different regimes of model shift.
First, we observe that our estimates accurately match with the empirical risks.
Second, as we vary the model shift parameter $\mu$, we observe the three information transfer regimes shown in Claim \ref{claim_model_shift}.

>>>>

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, and $\frac{p}{n_1+n_2} \le \frac 1 {2 + 2\tau}$).
Theorem \ref{thm_main_RMT} generalizes the classical result from Lemma \ref{fact_tr} to the sum of two independent sample covariance matrices.
To see this, we note that Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ (of the matrix $M$) are all equal to one, and $a = 1$.
In this setting, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above equation and the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the solutions of $\alpha_1, \alpha_2$ back to the LHS of equation \eqref{lem_cov_shift_eq}, we thus find that
\begin{align}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since $\lambda_i = 1$ for all $i=1,\dots,p$} \nonumber \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying the solutions of $\alpha_1$ and $\alpha_2$} \nonumber \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p}, \label{g_id}
\end{align}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the rate in Lemma \ref{fact_tr} for this particular setting.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.

@@

Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following result regarding $g(\id_{p \times p})$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2 n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        \bigtr{(\alpha_1 M^{\top} M + \alpha_2)^{-1}} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \ge \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
As a corollary of Claim \ref{prop_covariate}, when $n_1$ is much larger than $n_2 / \tau^2$, the excess risk of HPS is approximately optimal in the set of all possible $M\in\cS$ when there is no covariate shift between dataset one and two.

@@

\paragraph{Illustrative examples.}
While the estimates from Theorem \ref{cor_MTL_loss} are generally complex, we describe a simplified result for the random-effect model (cf. Section \ref{sec_data}).
Under the assumptions stated in Proposition \ref{lem_hat_v}, $\hat{a}$ is approximately equal to one.
Then, we can calculate that
\[ \cL_1(1) = \frac{p}{n_1 + n_2 - p} \text{ and } \cL_2(1) = \frac{n_1^2(n_1 + n_2 - p) + p n_1 n_2}{(n_1 + n_2)^2 (n_1 + n_2 - p)}, \text{ in which } \kappa(1) = \frac{n_2^2}{(n_1 + n_2)^2 - n_1 p}. \]
Thus, using that $\|\beta^{(1)}-\beta^{(2)}\|^2 = (2+\oo(1))\mu^2$ w.h.p., we conclude that for the random-effect model, the excess risk $L(\hat{\beta}_2^{\MTL})$ is approximately equal to
$$ g(n_1) \define \sigma^2 \cal L_1(1) +   2\mu^2 \cal L_2(1)= \frac{p\sigma^2}{n_1+n_2-p} +  2\mu^2 \cdot \frac{n_1^2 (n_1+n_2-p)+p n_1n_2}{(n_1+n_2)^2(n_1+n_2-p)}$$
plus lower order terms (that vanishes to zero as $p$ goes to infinity).
Next, we analyze when combining task one using HPS transfers positively to task two, depending on sample sizes $n_1, n_2$, and the model shift parameter $\mu$ in the random-effect model.

@@

As a remark, when $p < n_2 < 3p$, similar results can be derived using the above arguments (details omitted).
Figure \ref{fig_sec3_model_shift} illustrates Claim \ref{claim_model_shift} for different regimes of model shift parameter $\mu$ in the random-effect model.
This simulation affirms that our estimates accurately match with the empirical bias equation \eqref{Lbias} plus the variance equation \eqref{Lvar}.
Furthermore, we observe the three information transfer regimes shown in Claim \ref{claim_model_shift} by varying $\mu$ in the random-effect model.
